MATH 245 Homework 5

### 2024-04-09 PDEs/math245_hw5.Rmd
pde hw5 prob 2

---
title: "MATH 245 Homework 5"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-04-08"
output:
  pdf_document:
    extra_dependencies: ["tcolorbox", "xcolor"]
---

@@

@@

>>>>

---
title: "MATH 245 Homework 5"
author: "Ruby Krasnow and Tommy Thach"
date: "2024-04-09"
output:
  pdf_document:
    extra_dependencies: ["tcolorbox", "xcolor"]
---

@@

## Done: 2, 4, 5, 6
## Not started: 1, 3

@@

### Step 1

We will use the principle of superposition and write $u(t,x)=v(t,x)+w(t,x)$, so $v=u-w$. Then for $v(t,x)$, we have

\begin{equation}\label{eqn:inhomowaveV1}
    \begin{cases}
     \square v =F(t,x) - \square w = F(t,x)-w_{tt}+c^2 w_{xx} =H(t,x)    & 0<x<L, \quad t>0 \\
     v(0,x) = u(0,x)-w(0,x) = \phi(x) -w(0,x)  = \Phi(x)          & 0<x<L\\
     v_t(0,x) = u_t(0,x)-w_t(0,x)= \psi(x)  -  w_t(0,x)  = \Psi(x)      & 0<x<L\\
     v_x(t,0)=u_x(t,0)-w_x(t,0) = h(t)-h(t)=0 \\
     v_x(t,L)=u_x(t,L)-w_x(t,L) = g(t)-g(t)=0 & t>0
    \end{cases}       
\end{equation}

\begin{equation}\label{eqn:inhomowaveV}
    \begin{cases}
     \square v =H(t,x)   & 0<x<L, \quad t>0 \\
     v(0,x) =  \Phi(x)          & 0<x<L\\
     v_t(0,x) =\Psi(x)      & 0<x<L\\
     v_x(t,0)=0 \\
     v_x(t,L)=0 & t>0
    \end{cases}       
\end{equation}

and for $w$, we have
\begin{equation}\label{eqn:inhomowaveW}
    \begin{cases}
     w_{xx}=0 \\
     w_x(t,0)=h(t) \\
     w_x(t,L)=g(t)
    \end{cases}       
\end{equation}

For $w(t,x)$, we will use a function of the form 
$$w(t,x)= \left(x-\frac{x^2}{2L}\right)h(t)+\frac{x^2}{2L}g(t)$$
$$ w_{x}=h(t)- \frac{x}{L} h(t) + \frac{x}{L}g(t)$$
$$ w_{xx}=\frac{1}{L} \Bigl( g(t)- h(t) \Bigr) $$
So that we will have $w_x(t,0)= h(t), w_x(t,L)=g(t)$.

### Step 2:  Solve (\ref{eqn:inhomowaveV})

Assume (\ref{eqn:inhomowaveV}) has the separated solution $v(t,x)=X(x)T(t)$ and consider the following eigenvalue problem

\begin{equation*}
    \begin{cases}
     x''+\lambda X = 0 \\
     X'(0)=X'(L)=0
    \end{cases}       
\end{equation*}

W know that the solution to this eigenvalue problem is
$$\lambda_n= \left(\frac{n\pi}{L}\right)^2, \quad X_n(x)=\cos{\left( \frac{n \pi x}{L}\right)},  \quad n=0,1,2,3... $$

### Step 3

Now, let's write $v(t,x)$ as

$$v(t,x)= a_0(t)+\sum_{n=1}^{\infty}  a_n(t) \cos{  \left(\frac{n\pi x}{L}\right) } $$
and substitute this into the PDE of (\ref{eqn:inhomowaveV}).

$$v_{tt}(t,x)= a_0''(t)+\sum_{n=1}^{\infty}  a_n''(t) \cos{  \left(\frac{n\pi x}{L}\right) } $$
$$c^2 v_{xx}(t,x)= -\sum_{n=1}^{\infty} \left(\frac{n\pi c}{L}\right)^2 a_n(t) \cos{  \left(\frac{n\pi x}{L}\right) } $$

\begin{equation}\label{eqn:step3}
H(t,x)=a_0''(t)+ \sum_{n=1}^{\infty}\left[  a_n''(t) +  \left(\frac{n\pi c}{L}\right)^2 a_n(t)    \right] \cos{  \left(\frac{n\pi x}{L}\right) }
\end{equation}

### Step 4

For a fixed $t$, we write $H(t,x)$ as the Fourier cosine series
\begin{equation}\label{eqn:step4}
H(t,x)=q_0(t)+ \sum_{n=1}^{\infty}q_n(t) \cos{  \left(\frac{n\pi x}{l}\right) } \end{equation}

Such that
$$q_n (t)= \frac{2}{L} \int_{0}^{L}H(t,x)\cos{  \left(\frac{n\pi x}{L}\right) }dx $$
$$q_0(t)=\frac{1}{L} \int_{0}^{L}H(t,x)dx$$

### Step 5

\begin{tcolorbox}[colback=white, title=Variation of Parameters]
Consider an ODE $$(\ast)\quad y''+a(x)y'+b(x)y=c(x) $$
and assume $\{ y_1(x), y_2(x)\}$ are fundamental solutions of the associated homogeneous equation $$(\ast \ast)\quad y''+a(x)y'+b(x)y=0 $$
Then the general solution of $(\ast \ast)$ is 
$y_h(x)=c_1y_1(x)+c_2y_2(x)$ for $c_1, c_2 \in \mathbb{R}$ and the solution of $(\ast)$ is the sum of the general and particular solutions, $y=y_h+y_p$, where
$$y_p = -y_1(x) \int \frac{y_2(s)c(s)}{W(y_1,y_2)}ds+ y_2(x) \int \frac{y_1(s)c(s)}{W(y_1,y_2)}ds$$
where $$W(y_1,y_2)= \begin{vmatrix}
y_1 & y_2\\
y_1' & y_2'
\end{vmatrix}$$ \end{tcolorbox}

Combine (\ref{eqn:step3}) and (\ref{eqn:step4}):

$$q_0(t)=a_0''(t)$$
$$q_n (t)=  a_n''(t) +  \left(\frac{n\pi c}{L}\right)^2 a_n(t)   $$
First we find the fundamental solution of 
$$ a_0''(t) =0$$
$$a_0^H=A_0t+B_0$$
for $A_0, B_0 \in \mathbb{R}$. Call $y_1=1, y_2=t$.

$$W(1,t)= \begin{vmatrix}
1 & t\\
0 & 1 \end{vmatrix}= 1 $$

$$a_0^P(t) = -\int_0^t s \,q_n(s)\,ds+ t \int_0^t q_n(s)\,ds=\int_0^t(t-s)q_n(s)\,ds$$

$$a_0(t)=A_0t+B_0+\int_0^t(t-s)q_n(s)\,ds$$

Now we find the fundamental solutions of

$$ a_n''(t) +  \left(\frac{n\pi c}{L}\right)^2 a_n(t) =0$$
This has the characteristic equation
$r^2+  \left(\frac{n\pi c}{L}\right)^2 =0$, giving us $\{y_1,y_2\}=\left\{ \cos{\left(\frac{n\pi ct}{L}\right)}, \sin{\left(\frac{n\pi ct}{L}\right)} \right\}$.
Thus, the general solution to the homogeneous problem is

$$a_n^H(t)=A_n\cos{\left(\frac{n\pi ct}{L}\right)}+B_n\sin{\left(\frac{n\pi ct}{L}\right)}$$
for $A_n, B_n \in \mathbb{R}, n \in \mathbb{N}$.

$$W(y_1,y_2)(s)= \begin{vmatrix}
\cos{\left(\frac{n\pi cs}{L}\right)} & \sin{\left(\frac{n\pi cs}{L}\right)}\\
-\frac{n\pi c}{L}\sin{\left(\frac{n\pi cs}{L}\right)} & \frac{n\pi c}{L}\cos{\left(\frac{n\pi cs}{L}\right)}
\end{vmatrix}= \frac{n\pi c}{L}$$

\begin{align*}
a_n^P(t)& = -\cos{\left(\frac{n\pi ct}{L}\right)} \int_0^t \frac{\sin{\left(\frac{n\pi cs}{L}\right)}q_n(s)}{W(y_1,y_2)}ds + \sin{\left(\frac{n\pi ct}{L}\right)} \int_0^t \frac{\cos{\left(\frac{n\pi cs}{L}\right)}q_n(s)}{W(y_1,y_2)}ds \\
& =\int_0^t\frac{\sin{\left(\frac{n\pi c}{L} (t-s)\right)}q_n(s)}{W(y_1,y_2)}ds \\ &=\int_0^t\frac{\sin{\left(\frac{n\pi c}{L} (t-s)\right)}q_n(s)}{\frac{n\pi c}{L}}ds
\end{align*}

Therefore, $a_n(t)= a_n^H(t)+a_n^P(t)$, but the $A_n, B_n$ in $a_n^H(t)$ and the $A_0, B_0$ in $a_0^H(t)$ are still unknown.

### Step 6

To find $A_0, B_0, A_n, B_n$, we use the initial conditions from (\ref{eqn:inhomowaveV}).

$$\Phi(x)=v(0,x)= a_0(0)+\sum_{n=1}^{\infty}  a_n(0) \cos{  \left(\frac{n\pi x}{L}\right) } =B_0+\sum_{n=1}^{\infty}  A_n \cos{  \left(\frac{n\pi x}{L}\right) }$$

$$\Psi(x)=v_t(0,x)= A_0+\sum_{n=1}^{\infty} \frac{n\pi c}{L} B_n \cos{  \left(\frac{n\pi x}{L}\right) }$$

$$A_0= \frac{1}{L} \int_{0}^{L}\Psi(x)dx$$
$$B_0= \frac{1}{L} \int_{0}^{L}\Phi(x)dx$$
$$A_n= \frac{2}{L} \int_{0}^{L}\Phi(x)\cos{  \left(\frac{n\pi x}{L}\right) }dx$$
$$ \frac{n\pi c}{L}B_n= \frac{2}{L} \int_{0}^{L}\Psi(x)\cos{  \left(\frac{n\pi x}{L}\right) }dx$$

### Step 7: Solution

The solution of (\ref{eqn:inhomowave}) is then
$$u(t,x)= a_0(t)+\sum_{n=1}^{\infty}  a_n(t) \cos{  \left(\frac{n\pi x}{L}\right) }$$
where $$a_0(t)=\frac{1}{L} \int_{0}^{L} \left[t\,\Psi(x)+\Phi(x) \right]dx+\int_0^t(t-s)q_n(s)\,ds$$

$$a_n(t)=A_n\cos{\left(\frac{n\pi ct}{L}\right)}+B_n\sin{\left(\frac{n\pi ct}{L}\right)}+\int_0^t\frac{\sin{\left(\frac{n\pi c}{L} (t-s)\right)}q_n(s)}{\frac{n\pi c}{L}}ds$$
$$A_n= \frac{2}{L} \int_{0}^{L}\Phi(x)\cos{  \left(\frac{n\pi x}{L}\right) }dx, \quad B_n=\frac{2}{n\pi c}\int_{0}^{L}\Psi(x)\cos{  \left(\frac{n\pi x}{L}\right) }dx$$
$$q_n (t)= \frac{2}{L} \int_{0}^{L}H(t,x)\cos{  \left(\frac{n\pi x}{L}\right) }dx$$
\begin{align*}
H(t,x)&=F(t,x)-w_{tt}+c^2w_{xx} \\
&=F(t,x)-\left[\left(x-\frac{x^2}{2L}\right)h''(t)+\frac{x^2}{2L}g''(t)\right]
+\frac{c^2}{L} \Bigl( g(t)- h(t) \Bigr) \\
\Phi(x)  &= u(0,x)-w(0,x) \\
&= \phi(x) -w(0,x)  \\
&= \phi(x) -\left(x-\frac{x^2}{2L}\right)h(0)-\frac{x^2}{2L}g(0) \\
\Psi(x) &=u_t(0,x)-w_t(0,x) \\
&= \psi(x)  -  w_t(0,x)  \\
&= \psi(x)- \left(x-\frac{x^2}{2L}\right)h'(0)-\frac{x^2}{2L}g'(0)
\end{align*}
